Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-25 paper.tex
Move S3 GEL stuff to own section

Finally, to explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for chromosome 2 to an AWS S3
standard bucket. We repeated the benchmarks above on the same subset
of 562,640 variants, using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask
required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchrous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s
occurs at 64 processes, with a corresponding
network data rate of 121 MiB/s
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. Genotype data can therefore be fetched
from S3 around an order of magnitude more quickly than it can be decompressed
into RAM. This has some interesting consequences
when we consider the goal of seeking the highest possible compression levels
for genotype data and the usual cloud storage cost structuring models (see
Discussion).

@@

>>>>

@@

\subsection{Extensibility: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for the Genomics England
data chromosome 2 described in Table~\ref{tab-genomics-england-data}.
an AWS S3 standard bucket. 
[TODO clarify what we're doing and remind reader]
We repeated the benchmarks above on the same subset
of 562,640 variants, using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask
required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchrous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s
occurs at 64 processes, with a corresponding
network data rate of 121 MiB/s
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. Genotype data can therefore be fetched
from S3 around an order of magnitude more quickly than it can be decompressed
into RAM. [TODO QUESTION: can such rates be sustained, and be expected to 
happen in parallel? Do Durner et al consider this?]

% Maybe don't bother with this, the same trade-off always exists.
% but I guess it's an extra factor to take into account when weighing
% up practical engineering decisions.
% This has some interesting consequences
% when we consider the goal of seeking the highest possible compression levels
% for genotype data and the usual cloud storage cost structuring models (see
% Discussion).
